\mythesistitleNOTSYNPOSIS

### 2012-02-27 src/020chap2.tex
proof read TILL Sec 3.2.2

This chapter surveys several results that are significant to this
thesis or to \COP in general. These predominantly pertain to
characterizations of \COP, algorithmic tests to check for \COP,
optimization problems on binary matrices that do not have \COP and
some applications of \COP.

@@

\COP is closely connected to several types of graphs by way of
describing certain combinatorial graph properties. There are also
certain graphs, like convex bipartite graphs, that are defined solely
by one of its associated matrix having \COP.  In this section we will
see the relevance of \cop to graphs.  To see this we introduce certain
binary matrices that are used to define graphs in different
ways. While adjacency matrix is perhaps the most commonly used such
matrix, Definition~\ref{def:graphmatrices} defines this and a few
more.

\begin{definition}
  {\em Matrices that define graphs. \cite[Def.~2.4]{d08phd}.}  
  \label{def:graphmatrices} 
  Let $G$ and $H$ be defined as follows. $G = (V,E_G)$ is a graph with
  vertex set $V = \{v_i \mid i \in [n]\}$ and edge set $E_G \subseteq
  \{(v_i,v_j) \mid i, j \in [n]\}$ such that $|E_G| = m$. $H = (A, B,
  E_H)$ is a bipartite graph with partitions $A = \{a_i \mid i \in
  [n_a]\}$ and $B = \{b_i \mid i \in [n_b]\}$.
  \begin{enumerate}[{\ref{def:graphmatrices}}--i.]
  \item \emph{Adjacency matrix} of $G$ is the symmetric $n \times n$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(v_i,v_j) \in E_G$
    for all $i,j \in [n]$.
  \item \emph{Augmented adjacency matrix} of $G$ is obtained from its
    adjacency matrix by setting all main diagonal elements to \un, \ie
    $m_{i,i} = \un$ for all $i \in [n]$.
  \item \emph{Maximal clique matrix} or \emph{vertex-clique incidence
      matrix} of $G$ is the $n \times k$ binary matrix $M$ with
    $m_{i,j} = \un$ \iff $v_i \in C_j$ for all $i \in [n], j \in [k]$
    where $\{C_j \mid j \in [k]\}$ is the set of maximal cliques of
    $G$.
    \label{def::maxcliquematrix}
  \item \emph{Half adjacency matrix} of $H$ is the $n_a \times n_b$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(a_i, b_j) \in E_H$.
  \end{enumerate}
\end{definition}

@@

\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.
  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \cite{bhy00}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \cite{sb59}.  $G$ is an interval graph \iff its maximal clique matrix
    has COP \cite{fg65} (This follows \cite{gh64} which states
      that the maximal cliques of interval graph $G$ can be linearly
      ordered \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.)
    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.
    \end{enumerate}
    The set of unit interval graphs and the set of proper interval
    graphs coincide \cite{rob69, gar07}.
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.
  \item $H$ is \emph{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

@@

Without doubt, besides computing answers to these questions, we are
interested in the efficiency of these computations in terms of
computational complexity theory. Results towards
questions~(\ref{q:testcop}) and (\ref{q:copperm}) are surveyed in this
section. Those for question~(\ref{q:copopt}) are discussed in
Section~\ref{sec:surveycopopt} and question~(\ref{q:nocopcert}) is discussed
in Section~\ref{sec:surveycertalgo}.

It may be noted that one way to design an algorithm to test for \COP
is by deriving one from any interval graph recognition algorithm using
the result \cite{hmpv00} \cite[Th~2.7]{d08phd} which demonstrates how
such a derivation can be done.  They use something called {\em Lex-BFS
  ordering} of the vertices of a graph to decide in linear time
whether the graph is an interval graph. They also show how any
interval graph recognition algorithm can be used to recognize matrices
with \COP. This is shown in Theorem~\ref{th:intgraphcop}.

@@

However, this does not necessarily yield an efficient algorithm. We
will see results that directly solve the problem on matrices since it
is known that questions~(\ref{q:testcop}) and (\ref{q:copperm}) stated
above for \COP are efficiently solvable.  Table~\ref{tab:cophistory}
gives a snapshot of these results.

@@

\begin{definition}[Asteroidal triple]
  \label{def:asteroidal}
  % If $G = (V,E)$ is a graph, a set of three vertices from $V$ form
  % an {\em asteroidal triple} if between any two of them there exists
  % a path in $G$ that does not contain any vertex from the closed
  % neighborhood of the third vertex.
  An {\em asteroidal triple} in a graph $G$ is a tuple of vertices
  $\{v_1,v_2,v_3\} \subset V$ such that between every pair of vertices
  from this tuple there exists a path that avoids the closed
  neighbourhood of the third vertex.
\end{definition}

@@

%\begin{minipage}[h]{5.5in}
\begin{definition}[\PQtree \cite{bl76, mcc04}]
  A \PQtree of matrix $M$ with \COP on columns (rows), is a tree with
  the following properties.
  \begin{enumerate}[i.]
    \singlespacing
  \item Each leaf uniquely represents a row (column) of $M$. The leaf
    order of the tree gives a \COP order for column (row) (\COP order
    for column requires permutation of rows and vice versa.) for $M$.
  \item Every non-leaf node in the tree is labeled $P$ or $Q$.
  \item \label{def::nodep} The children of $P$ nodes are
    unordered. They can be permuted in any fashion to obtain a new
    \COP order for $M$.
  \item \label{def::nodeq} The children of $Q$ nodes are linearly
    ordered. Their order can be reversed to obtain a new \COP
    order for $M$.
  \end{enumerate}
  \label{def:pqtree}
\end{definition}
%\end{minipage}

@@

Judging from notes in literature, this algorithm is apparently
notoriously difficult to program. In the procedure to modify the
\PQtree at each iteration, nodes are considered from leaves to
tree. At each node considered, it uses one of nine templates to
determine how the tree must be altered in the vicinity of this
node. Recognition of this template poses a difficult challenge in
terms of implementing it. Each template is actually a representative
of a larger class of similar templates, which must be dealt with
explicitly by a program\cite{mcc04}.

@@

\begin{definition}[{\em \Cop for set systems}]%
  \label{def:matrixsetsystem}%
  % [\emph{Set system of a binary matrix}.]
  Let $M$ be a binary matrix of order $n \times m$ and $\{c_i \mid i
  \in [m]\}$ be the columns in $M$.  A set system $\cF_{M} = \{S_i
  \mid S_i \subseteq [n], i \in [m]\}$ is defined such that for every
  column $c_i$ of $M$, set $S_i = \{j \mid m_{ji} = \un \}$. The
  collection $\cF_M$ is the {\em set system of binary matrix} $M$. The
  {\em binary matrix for set system} $\cF$ is conversely constructed
  and denoted by $M^\cF$. Thus, $M^{\cF_M} = M$.%
  \par\noindent%
  A set system $\cF$ from universe $U$, $|U| = n$ has the {\em \cop}
  if there exists a linear order or permutation $\sigma = w_1w_2\ldots
  w_n$ that can be applied to $U$ \stt each set $S \in \cF$ becomes a
  consecutive subsequence (also termed an {\em interval})
  $w_{i}w_{i+1}\ldots w_{i+k-1}$ on $\sigma$ for some positive integer
  $i \le n+1-k$ where $k = |S|$.
  \par\noindent%
  The set $\valid{U,\cF}$ represents all \COP orders of $\cF$ in $U$.
  \dstop
\end{definition}

@@

\begin{definition}[\em Orthogonal sets \cite{n89, mm96,
    mcc04}\footnote{\cite{mcc04} does not use the term ``mutually
    orthogonal'' but refers to the same idea as ``sets that do not
    overlap''. This terminology is also used in other literature like
    \cite{nsnrs09,wlh02}.} etc.]
  \label{def:orthosets}%
  Let $\cF$ be a set system with universe $U$ and sets $A, B \in \cF$.
  \begin{enumerate}
  \item $A$ and $B$ are said to have a {\em trivial intersection} if
    $A \cap B$ is $\emptyset$ or $A$ or $B$.  In other words, $A$ and
    $B$ are either disjoint or one is the subset of the other.

  \item $A$ and $B$ are called {\em mutually orthogonal} or $A$ is {\em
      orthogonal to} $B$ and vice versa, if they have a trivial
    intersection.

  \item {\em Trivial subsets} of a universe $U$, denoted by $\cT(U)$,
    are sets that have trivial intersections with any set in
    $\power{U}$. These sets are $U$, singleton sets in
    $\power{U}$ and $\emptyset$ \cite{n89, mm96}.
    Thus, $\cT(U) = \{U\} \  \bigcup \  \{ \{v\} \mid v \in U \} \  \bigcup \
    \{\emptyset\}$.

  \item {\em Orthogonal sets of a set
      system}\footnote{\cite[Def.~3.1]{mcc04} uses the term {\em
        non-overlapping family} of $\cF$ and denotes it by
      $\cN(\cF)$. \label{mcc1}} $\cF$ with universe $U$ are subsets of
    $U$ that are orthogonal to all sets in $\cF$. The set of all
    orthogonal sets to $\cF$ is denoted by $\cF^{\bot}$.

  \item $\cF$ is called {\em complete}\footnote{\cite{mcc04} calls
      this a {\em weakly partitive family}. \label{mcc2}} if the
    following hold true for every pair of non-orthogonal (overlapping)
    sets $A, B$ in $\cF$.

@@

  \item $\overline{\cF}$ represents the smallest super set system of
    $\cF$ that is complete\footnote{\cite[Def.~3.2]{mcc04} calls
      this the {\em weak closure} of $\cF$ denoted by
      $\cW(\cF)$. \label{mcc3}}

@@

\begin{observation}[{\cite[Sec.~3]{mm96}}]
  If $\cF$ is a set system with \COP then, after applying the \COP
  order, not only must every set in $\cF$ be consecutive but the
  following sets must also be consecutive for any $A, B \in \cF$.
\begin{enumerate}
  \item The intersection $A \cap B$ 
  \item The union $A \cup B$ if $A \cap B \ne \emptyset$
  \item The relative complements $A \setminus B$ and $B \setminus A$
    if $B \nsubseteq A$ and $A \nsubseteq B$ respectively.
  \item Also note that trivially, sets in $\cT(U)$ are consecutive in
    any permutation of $U$\footnote{$\emptyset$ is considered
      consecutive by convention.}.
\end{enumerate}
\end{observation}

The following theorem gives a very interesting property of orthogonal
sets and $\overline{\cF}$

@@

The following corollary, Corollary~\ref{cor:pqrrecursive} states how
\PQRtree elegantly fits into this whole theory and help in computing
all \COP orders of $\cF$.

\begin{corollary}[{\cite[Cor.~8]{mm96}}]
  \label{cor:pqrrecursive}
  Let $\cF$ be a set system with universe $U$ and $H$ is a non-empty
  orthogonal set $H \in \ff$. If there is a \PQRtree $T_1$ that
  encodes all permutations in $\valid{U/H, \cF/H}$ and also a \PQRtree
  $T_2$ that encodes all permutations $\valid{H,\cF \cap \power{H}}$,
  then a \PQRtree $T$ for $\cF$ can be obtained by replacing the leaf
  $h$ in $T_1$ by $T_2$, where $h$ is the representative element for $H$.
\end{corollary}

@@

\begin{theoremsansproof}[{\cite[Th.~9]{mm96}}] %,\cite[Th.~2.1]{mcc04}
\label{th:pqrfinalmm}
If $\cF$ is a set system with universe $U$ and $|U| \ge 3$ then one of
the following statements hold.
\begin{enumerate}
\item \label{th::it1} $\overline{\cF} = \cT(U)$ 
\item \label{th::it2}$\overline{\cF} = consec(\alpha)$ for
  some permutation $\alpha$ on $U$ ($consec(\alpha)$ is a linear order)
\item \label{th::it3}$\overline{\cF} = \power{U}$ 
\end{enumerate}
\end{theoremsansproof}

@@

Going back to the theory we were discussing earlier,
Theorem~\ref{th:pqrfinalmm} and the theory leading to it is very
similar to \cite[esp. Th.~2.1, 3.5. also Th.~3.2, 3.3, 3.4]{mcc04}
which categorizes the nodes in the above the three cases as {\em
  prime}, {\em linear} and {\em degenerate}
respectively. \cite{mcc04}'s generalized \PQtree is created in similar
ways as \PQRtree above. This tree is, in essence, the Hasse diagram of
what they call {\em strong elements} of weak closure, $\cW(\cF)$(\ie
$\overline{\cF}$). Strong elements of a set family are elements that
do not overlap with any other elements in the family, \ie it is
orthogonal to all other sets\cite[Def.~3.3]{mcc04}.

@@

We will now see, in brief, one more generalization of \PQtree before
seeing another approaches to solving \COP testing called ICPIA which
is a set theoretic characterization of \COP.

@@

    &
    \parbox[t]{\colwidth}
    {
      $P$-node $\Leftrightarrow \cF = \cT(\cF)$\\
      $Q$-node $\Leftrightarrow \cF = consec(\alpha)$\\
      $R$-node $\Leftrightarrow \cF = \power{U}$\\
    }
    &
    \parbox[t]{\colwidth}
    {
      {\em Prime}: For every $I$, $\emptyset \ne I \subset [k],
      \cup_{i \in I}S_i \notin \cF$\\
      {\em Linear}: There exists an ordering of $[k]$
      s.t. $\emptyset \ne I \subset [k]$ implies $\cup_{i \in I}S_i
      \in \cF$ only if members of $I$ are consecutive in the ordering\\
      {\em Degenerate}: For every $I$, $\emptyset \ne I \subset [k],
      \cup_{i \in I}S_i \in \cF$\\
    }

@@

To see it in terms of the theory explained in
Section~\ref{sec:surveycertalgo}, this results in the weak closure of
$\cF$. Let $\cF$, $\cI$ be the set system and interval
system (A set system where all the sets are intervals.),
respectively, involved in the assignment $\Pi$ and $\cF'$, $\cI'$ be
the same for the output of the algorithm $\Pi'$.  It can be observed
that $\cF' \cup \cF$ is $\overline{\cF}$ or the weak closure of
$\cF$. The analog is true for $\cI \cup \cI'$.  Moreover, this output
$\cF'$, $\cI'$ contains the strong elements of $\overline{\cF}$,
$\overline{\cI}$ respectively. It turns out that $\cI'$ is also an
interval system and that $\Pi'$ is an ICPIA \cite[Lem.~2]{nsnrs09}.

The next algorithm \cite[Alg.~2]{nsnrs09} further refines $\Pi'$ to
$\Pi''$ which represents the family of permutations that yield the
\COP orders represented by $\Pi$ of $\cF$. The subset relation Hasse
diagram of $I'$ is created with each node representing the
corresponding assignment pair of the interval from $\Pi'$. Since all intervals in
$\Pi'$ are non-overlapping, this is a tree. Notice that this means any
two intervals are now either disjoint or one is contained in the
other. Hence this tree is called a {\em containment tree}.  The
algorithm traverses this tree in post order fashion where the post
order function is to replace an assignment pair $(X,Y)$  in $\Pi'$ by their child
assignment pairs in the containment tree, say $(X_1, Y_1),(X_2, Y_2)
\ldots (X_k,Y_k)$. The result $\Pi''$ gives the
\COP order for $\Pi$. These two algorithms prove that ICPIA is a
sufficient condition for the feasibility of the interval assignment.

@@

It is known from Theorem~\ref{th:tuckercop} that when a matrix does
not have \COP, it must have one of Tucker's forbidden submatrices (See
Section~\ref{sec:tucker} for detailed discussion on Tucker's forbidden
substructures for matrices with \COP). However, none of the \COP test
algorithms discussed in Section~\ref{sec:surveycoptest} outputs the
forbidden submatrix when the test is negative. This perhaps gives a
hint about how hard it is to find these substructures in the input
matrix. Using a reduction to the longest path problem, \cite{v85}
proves that finding a Tucker's forbidden submatrix of at least $k$
rows is NP complete. An example of an optimization problem for a
matrix without \COP is finding if deleting at most $k$ rows from input
will restore \COP. This is known to be NP complete
\cite{b75-phd}. There are several such optimization problems that try
to find ways to get \COP on the input matrix with certain alterations
and finding smallest forbidden submatrix is at the heart of these problems.

In this section we will summarize \cite{d08phd}'s approximation
algorithm for finding forbidden submatrix. We will use $\Tuckerset$ to
refer to the set of Tucker's forbidden submatrices shown in
Figure~\ref{fig:forbiddensubmatrices}.

>>>>

This chapter surveys several results that are significant to this
thesis or to \COP in general. These predominantly pertain to
characterizations of \COP, algorithmic tests to check for \COP
and certificates for binary matrices that do not have \COP.

@@

\COP is closely connected to several types of graphs by way of
describing certain combinatorial graph properties. There are also
certain graphs, like convex bipartite graphs, that are defined solely
by one of its associated matrices having \COP.  In this section we will
see the relevance of \cop to graphs.  To see this we introduce certain
binary matrices that are used to define graphs in different
ways. While adjacency matrix is perhaps the most commonly used such
matrix, Definition~\ref{def:graphmatrices} defines this and a few
more.

\begin{definition}
  [Matrices that define graphs {\cite[Def.~2.4]{d08phd}}]  
  \label{def:graphmatrices} 
  Let $G$ and $H$ be defined as follows. $G = (V,E_G)$ is a graph with
  vertex set $V = \{v_i \mid i \in [n]\}$ and edge set $E_G \subseteq
  \{(v_i,v_j) \mid i, j \in [n]\}$ such that $|E_G| = m$. $H = (A, B,
  E_H)$ is a bipartite graph with partitions $A = \{a_i \mid i \in
  [n_a]\}$ and $B = \{b_i \mid i \in [n_b]\}$.
  \begin{enumerate}[{\ref{def:graphmatrices}}--i.]
  \item \textbf{Adjacency matrix} of $G$ is the symmetric $n \times n$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(v_i,v_j) \in E_G$
    for all $i,j \in [n]$.
  \item \textbf{Augmented adjacency matrix} of $G$ is obtained from its
    adjacency matrix by setting all main diagonal elements to \un, \ie
    $m_{i,i} = \un$ for all $i \in [n]$.
  \item \textbf{Maximal clique matrix} or \textbf{vertex-clique incidence
      matrix} of $G$ is the $n \times k$ binary matrix $M$ with
    $m_{i,j} = \un$ \iff $v_i \in C_j$ for all $i \in [n], j \in [k]$
    where $\{C_j \mid j \in [k]\}$ is the set of maximal cliques of
    $G$.
    \label{def::maxcliquematrix}
  \item \textbf{Half adjacency matrix} of $H$ is the $n_a \times n_b$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(a_i, b_j) \in E_H$.
  \end{enumerate}
\end{definition}

@@

\begin{definition}[Graphs that relate to
  COP {\cite[Def.~2.5]{d08phd}}] %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.
  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \textbf{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \textbf{concave-round} if its
    augmented adjacency matrix has \CROP. \cite{bhy00}
  \item $G$ is an \textbf{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \cite{sb59}.  $G$ is an interval graph \iff its maximal clique matrix
    has COP \cite{fg65} (This follows \cite{gh64} which states
      that the maximal cliques of interval graph $G$ can be linearly
      ordered \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.)
    \begin{enumerate}[a.]
    \item $G$ is a \textbf{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.
    \item $G$ is a \textbf{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.
    \end{enumerate}
    The set of unit interval graphs and the set of proper interval
    graphs coincide \cite{rob69, gar07}.
 \item $G$ is a \textbf{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.
  \item $H$ is \textbf{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \textbf{biconvex bipartite} or \textbf{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \textbf{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

@@

Without doubt, besides computing answers to these questions, we are
interested in the efficiency of these computations in terms of
computational complexity theory. Results towards
questions~(\ref{q:testcop}) and (\ref{q:copperm}) are surveyed in this
section. Those for question~(\ref{q:nocopcert}) is discussed in
Section~\ref{sec:surveycopopt}. Question~(\ref{q:copopt}) is discussed
in brief in Section~\ref{sec:optcop} -- it is not elaborated in this
chapter since the focus of this thesis is characterization and
recognition rather than optimization.

One way to design an algorithm to test for \COP is by deriving one
from any interval graph recognition algorithm using the result
\cite{hmpv00} \cite[Th~2.7]{d08phd}, restated below in
Theorem~\ref{th:intgraphcop}, which demonstrates how such a derivation
can be done.  They use something called {\em Lex-BFS ordering} of the
vertices of a graph to decide in linear time whether the graph is an
interval graph. They also show how any interval graph recognition
algorithm can be used to recognize matrices with \COP.

@@

However, this does not necessarily yield an efficient algorithm. We
will see results that directly solve the problem of \COT on matrices
since it is known that questions~(\ref{q:testcop}) and
(\ref{q:copperm}) stated above for \COP are efficiently solvable.
Table~\ref{tab:cophistory} gives a snapshot of these results.

@@

\begin{definition}[Asteroidal triple]
  \label{def:asteroidal}
  % If $G = (V,E)$ is a graph, a set of three vertices from $V$ form
  % an {\em asteroidal triple} if between any two of them there exists
  % a path in $G$ that does not contain any vertex from the closed
  % neighborhood of the third vertex.
  An \textbf{asteroidal triple} in a graph $G$ is a tuple of vertices
  $\{v_1,v_2,v_3\} \subset V$ such that between every pair of vertices
  from this tuple there exists a path that avoids the closed
  neighbourhood of the third vertex.
\end{definition}

@@

%\begin{minipage}[h]{5.5in}
\begin{definition} [\PQtree \cite{bl76, mcc04}]
  A \PQtree of matrix $M$ with \COP on columns (rows), is a tree with
  the following properties.
  \begin{enumerate}[i.]
    \singlespacing
  \item Each leaf uniquely represents a row (column) of $M$. The leaf
    order of the tree gives a \COP order for column (row) (\COP order
    for column requires permutation of rows and vice versa.) for $M$.
  \item Every non-leaf node in the tree is labeled $P$ or $Q$.
  \item \label{def::nodep} The children of $P$ nodes are
    unordered. They can be permuted in any fashion to obtain a new
    \COP order for $M$.
  \item \label{def::nodeq} The children of $Q$ nodes are linearly
    ordered. Their order can be reversed to obtain a new \COP
    order for $M$.
  \end{enumerate}
  \label{def:pqtree}
\end{definition}
%\end{minipage}

@@

Judging from notes in literature, this algorithm is apparently
notoriously difficult to program. In the procedure to modify the
\PQtree at each iteration, nodes are considered from leaves to
the root. At each node considered, it uses one of nine templates to
determine how the tree must be altered in the vicinity of this
node. Recognition of this template poses a difficult challenge in
terms of implementing it. Each template is actually a representative
of a larger class of similar templates, which must be dealt with
explicitly by a program\cite{mcc04}.

@@

\begin{definition}[\Cop for set systems]%
  \label{def:matrixsetsystem}%
  % [\emph{Set system of a binary matrix}.]
  Let $M$ be a binary matrix of order $n \times m$ and $\{c_i \mid i
  \in [m]\}$ be the columns in $M$.  A set system $\cF_{M} = \{S_i
  \mid S_i \subseteq [n], i \in [m]\}$ is defined such that for every
  column $c_i$ of $M$, set $S_i = \{j \mid m_{ji} = \un \}$. The
  collection $\cF_M$ is the \textbf{set system of binary matrix} $M$. The
  \textbf{binary matrix for set system} $\cF$ is conversely constructed
  and denoted by $M^\cF$. Thus, $M^{\cF_M} = M$.%
  \par\noindent%
  A set system $\cF$ from universe $U$, $|U| = n$ has the \textbf{\cop}
  if there exists a linear order or permutation $\sigma = w_1w_2\ldots
  w_n$ that can be applied to $U$ \stt each set $S \in \cF$ becomes a
  consecutive subsequence (also termed an \textbf{interval})
  $w_{i}w_{i+1}\ldots w_{i+k-1}$ on $\sigma$ for some positive integer
  $i \le n+1-k$ where $k = |S|$.
  \par\noindent%
  The set $\valid{U,\cF}$ represents all \COP orders of $\cF$ in $U$.
  \dstop
\end{definition}

@@

\begin{definition}[Orthogonal sets \cite{n89, mm96,
    mcc04}\footnote{\cite{mcc04} does not use the term ``mutually
    orthogonal'' but refers to the same idea as ``sets that do not
    overlap''. This terminology is also used in other literature like
    \cite{nsnrs09,wlh02}.} etc.]
  \label{def:orthosets}%
  Let $\cF$ be a set system with universe $U$ and sets $A, B \in \cF$.
  \begin{enumerate}
  \item $A$ and $B$ are said to have a \textbf{ trivial intersection} if
    $A \cap B$ is $\emptyset$ or $A$ or $B$.  In other words, $A$ and
    $B$ are either disjoint or one is the subset of the other.

  \item $A$ and $B$ are called \textbf{ mutually orthogonal} or $A$ is \textbf{
      orthogonal to} $B$ and vice versa, if they have a trivial
    intersection.

  \item \textbf{ Trivial subsets} of a universe $U$, denoted by $\cT(U)$,
    are sets that have trivial intersections with any set in
    $\power{U}$. These sets are $U$, singleton sets in
    $\power{U}$ and $\emptyset$ \cite{n89, mm96}.
    Thus, $\cT(U) = \{U\} \  \bigcup \  \{ \{v\} \mid v \in U \} \  \bigcup \
    \{\emptyset\}$.

  \item {\em Orthogonal sets of a set
      system}\footnote{\cite[Def.~3.1]{mcc04} uses the term \textbf{
        non-overlapping family} of $\cF$ and denotes it by
      $\cN(\cF)$. \label{mcc1}} $\cF$ with universe $U$ are subsets of
    $U$ that are orthogonal to all sets in $\cF$. The set of all
    orthogonal sets to $\cF$ is denoted by $\cF^{\bot}$.

  \item $\cF$ is called \textbf{ complete}\footnote{\cite{mcc04} calls
      this a \textbf{ weakly partitive family}. \label{mcc2}} if the
    following hold true for every pair of non-orthogonal (overlapping)
    sets $A, B$ in $\cF$.

@@

  \item $\overline{\cF}$ represents the smallest super set system of
    $\cF$ that is complete\footnote{\cite[Def.~3.2]{mcc04} calls
      this the \textbf{ weak closure} of $\cF$ denoted by
      $\cW(\cF)$. \label{mcc3}}

@@

\begin{observation}[{\cite[Sec.~3]{mm96}}]
  If $\cF$ is a set system with \COP then, after applying the \COP
  order, not only must every set in $\cF$ be consecutive but the
  following sets must also be consecutive for any $A, B \in \cF$.
\begin{enumerate}
  \item The intersection $A \cap B$ 
  \item The union $A \cup B$ if $A \cap B \ne \emptyset$
  \item The relative complements $A \setminus B$ and $B \setminus A$
    if $B \nsubseteq A$ and $A \nsubseteq B$ respectively.
  \item Also note that trivially, sets in $\cT(U)$ are consecutive in
    any permutation of $U$ ($\emptyset$ is considered
      consecutive by convention.).
\end{enumerate}
\end{observation}

The following theorem gives a very interesting property of orthogonal
sets and $\overline{\cF}$.

@@

The following Corollary~\ref{cor:pqrrecursive} states how
\PQRtree elegantly fits into this whole theory and help in computing
all \COP orders of $\cF$, if any.

\begin{corollary}[{\cite[Cor.~8]{mm96}}]
  \label{cor:pqrrecursive}
  Let $\cF$ be a set system with universe $U$ and $H$ is a non-empty
  orthogonal set $H \in \ff$. If there is a \PQRtree $T_1$ that
  encodes all permutations in $\valid{U/H, \cF/H}$ and also a \PQRtree
  $T_2$ that encodes all permutations $\valid{H,\cF \cap \power{H}}$,
  then a \PQRtree $T$ for $\cF$ can be obtained by replacing the leaf
  $h$ in $T_1$ by $T_2$, where $h$ is the representative element for
  $H$ in $T_i$.
\end{corollary}

@@

\begin{theoremsansproof}[{\cite[Th.~9]{mm96}}] %,\cite[Th.~2.1]{mcc04}
\label{th:pqrfinalmm}
If $\cF$ is a set system with universe $U$ and $|U| \ge 3$ then one of
the following statements hold.
\begin{enumerate}
\item \label{th::it1} $\overline{\cF} = \cT(U)$ 
\item \label{th::it2}$\overline{\cF} = consec(\alpha)$ for
  some permutation $\alpha$ on $U$ ($consec(\alpha)$ denotes a linear order)
\item \label{th::it3}$\overline{\cF} = \power{U}$ 
\end{enumerate}
\end{theoremsansproof}

@@

Going back to the theory we were discussing earlier,
Theorem~\ref{th:pqrfinalmm} and the theory leading to it is very
similar to \cite[esp. Th.~2.1, 3.5. also Th.~3.2, 3.3, 3.4]{mcc04}
which categorizes the nodes in the three cases of
Theorem~\ref{th:pqrfinalmm} as {\em prime}, {\em linear} and {\em
  degenerate} respectively. \cite{mcc04}'s generalized \PQtree is
created in similar ways as \PQRtree above. This tree is, in essence,
the Hasse diagram of what they call {\em strong elements} of weak
closure, $\cW(\cF)$(\ie $\overline{\cF}$). Strong elements of a set
family are elements that do not overlap with any other elements in the
family, \ie it is orthogonal to all other sets\cite[Def.~3.3]{mcc04}.

@@

We will now see, in brief, one more generalization of \PQtree before
seeing another approach to solving \COP testing called ICPIA which
is a set theoretic characterization of \COP.

@@

    &
    \parbox[t]{\colwidth}
    {
      $P$-node $\Leftrightarrow \cF = \cT(\cF)$\\
      $Q$-node $\Leftrightarrow \cF = consec(\alpha)$\\
      $R$-node $\Leftrightarrow \cF = \power{U}$\\
    }
    &
    \parbox[t]{\colwidth}
    {
      \textbf{Prime}: For every $I$, $\emptyset \ne I \subset [k],
      \cup_{i \in I}S_i \notin \cF$\\
      \textbf{Linear}: There exists an ordering of $[k]$
      s.t. $\emptyset \ne I \subset [k]$ implies $\cup_{i \in I}S_i
      \in \cF$ only if members of $I$ are consecutive in the ordering\\
      \textbf{Degenerate}: For every $I$, $\emptyset \ne I \subset [k],
      \cup_{i \in I}S_i \in \cF$\\
    }

@@

To see it in terms of the theory explained in
Section~\ref{sec:surveycertalgo}, this results in the set of strong
elements of the weak closure of
$\cF$. Let $\cF$, $\cI$ be the set system and interval
system (a set system where all the sets are intervals),
respectively, involved in the assignment $\Pi$ and $\cF'$, $\cI'$ be
the same for the output of the algorithm $\Pi'$.  It can be observed
that $\cF' \cup \cF$ is $\overline{\cF}$ or the weak closure of
$\cF$ without the set unions. The analog is true for $\cI \cup \cI'$.  Moreover, this output
$\cF'$, $\cI'$ contains the strong elements of $\overline{\cF}$,
$\overline{\cI}$ respectively. It turns out that $\cI'$ is also an
interval system and that $\Pi'$ is an ICPIA \cite[Lem.~2]{nsnrs09}.

The next algorithm \cite[Alg.~2]{nsnrs09} further refines $\Pi'$ to
$\Pi''$ which represents the family of permutations that yield the
\COP orders represented by $\Pi$ of $\cF$. The subset relation Hasse
diagram of $\cI'$ is created with each node representing the
corresponding assignment pair of the interval from $\Pi'$. Since all intervals in
$\Pi'$ are non-overlapping, this is a tree. Notice that this means any
two intervals are now either disjoint or one is contained in the
other. Hence this tree is called a {\em containment tree}.  The
algorithm traverses this tree in post order fashion where the post
order function is to replace an assignment pair $(X,Y)$  in $\Pi'$ by their child
assignment pairs in the containment tree, say $(X_1, Y_1),(X_2, Y_2)
\ldots (X_k,Y_k)$. The result $\Pi''$ gives the
\COP order for $\Pi$. These two algorithms prove that ICPIA is a
sufficient condition for the feasibility of the interval assignment.

@@

% When a \COP test algorithm reports negative, it is not easy to verify
% this. We already saw that \cite{at72} gives forbidden substructures
% for \COP. \cite{v82} shows that the maximization problem of finding a forbidden submatrix with at
% least $k$ rows in an input matrix is NP-complete. This was proven by
% polynomially reducing the longest path problem for graphs to this
% problem. For most cases, the minimization problem is more helpful.
% Section~\ref{sec:forbiddensubalgo} gave a polynomial
% algorithm for finding minimum size forbidden submatrix.

It is known from Theorem~\ref{th:tuckercop} that when a matrix does
not have \COP, it must have one of Tucker's forbidden submatrices (See
Section~\ref{sec:tucker} for detailed discussion on Tucker's forbidden
substructures for matrices with \COP). However, none of the \COP test
algorithms discussed in Section~\ref{sec:surveycoptest} outputs the
forbidden submatrix when the test is negative.  Using a reduction to
the longest path problem, \cite{v82} proves that finding a Tucker's
forbidden submatrix of at least $k$ rows is NP complete. This is a
maximization problem.  An example of an optimization problem for a
matrix without \COP is finding if deleting at most $k$ rows from input
will restore \COP. This is known to be NP complete
\cite{b75-phd}. There are several such optimization problems that try
to find ways to get \COP on the input matrix with certain alterations
and finding smallest forbidden submatrix is at the heart of these
problems. This minimization problem of finding smallest forbidden
submatrix is polynomial time solvable.
